---
title: "Gov 51 Problem Set 2"
author: "Caleb Masshardt"
format: pdf
editor: visual
---

GitHub URL: <https://github.com/cmasshardt/gov51-ps2.git>

## Part 1: The Original Study

**1.1 Exploring the Experiment**

*1.1.1 Load packages and data*

```{r}
#| include: false
library(tidyverse)
gay <- read_csv("./data/raw/gay.csv")
```

***Q1.1 (3 points):** How many total observations are in gay.csv? How many unique values does study take? List all the treatment conditions in the treatment variable.*

```{r}
#Number of observations
nrow(gay)

#Number of unique values "study" takes
table(gay$study)

#Listing treatment conditions in the "treatment" variable
table(gay$treatment)  
```

There are 69,592 observations (rows) in gay.csv.

The variable "study" takes two different unique values: 1 and 2.

The "treatment" variable takes on one of five conditions: "No contact," "Recycling script by gay canvasser," "Recycling script by straight canvasser," "Same-sex marriage script by gay canvasser," or "Same-sex marriage script by straight canvasser."

***Q1.2 (3 points):** Filter to Study 1, Wave 1 only. Calculate the mean of ssm (same-sex marriage attitude, scored 1–5) for each treatment group using group_by() and summarize(). Which treatment group shows the highest average support?*

```{r}
#Subsetting to study 1, wave 1
study1 <- gay[gay$study==1,]
s1w1 <- study1[study1$wave==1,]

#Finding mean of ssm by treatment group
mean_ssm_by_treatment <- s1w1 %>%
  group_by(treatment) %>%
  summarize(mean_ssm = mean(ssm))
mean_ssm_by_treatment

```

The treatment group "recycling script by gay canvasser" shows the highest average support, with a mean same-sex marriage attitude of 3.13.

***Q1.3 (3 points):** Still using Study 1, Wave 1, calculate the standard deviation of ssm within each treatment group. Are the standard deviations roughly similar across groups? Why is this important in a randomized experiment? (2–3 sentences)*

```{r}
#Finding standard deviations of ssm treatment group
sd_ssm_by_treatment <- s1w1 %>%
  group_by(treatment) %>%
  summarize(sd_ssm = sd(ssm))
sd_ssm_by_treatment
```

Yes, the standard deviations are roughly similar across groups, encompassing the limited range between 1.66 and 1.70. In a randomized experiment, it is important that the standard deviations are roughly similar across groups because this confirms that the groups are randomized well and do not contain systematic differences from one another. Since we are measuring the effect of a treatment on the response variable (ssm), the baseline response variable values need to be similar so that we can meaningfully compare the changes in each group that result from the treatment.

***Q1.4 (3 points):** Create a bar plot of the mean ssm by treatment group (Study 1, Wave 1) using ggplot() with geom_col(). Hint: First create a summary data frame with the means, then plot it.*

```{r}
#Reading in scales library to enable label wrap
library(scales)

#Making bar plot for mean SSM by treatment group 
ggplot(mean_ssm_by_treatment, aes(x=treatment, y=mean_ssm)) +
  geom_col(width=0.8, fill="steelblue") +
  labs(title = "Mean SSM by Treatment Group") +
  labs(x = "Treatment Group", y = "Mean SSM") +
  scale_x_discrete(labels = label_wrap(10)) +
  theme_minimal()
```

**1.2 Checking Balance and Comparing Outcomes**

***Q1.5 (4 points):** A well-run randomized experiment should produce groups that look similar at baseline. Filter gay.csv to Study 1, Wave 1. Compare the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” group:*

*• Calculate the mean and SD of ssm for each group*

*• Compute the standardized difference in means:*

*A standardized difference below 0.25 in absolute value suggests good balance. Do these groups look similar at baseline? (2–3 sentences)*

```{r}
#Mean ssm for "Same-Sex Marriage Script by Gay Canvasser"
ss_gay <- s1w1[s1w1$treatment=="Same-Sex Marriage Script by Gay Canvasser",]
mean(ss_gay$ssm)

#Standard deviation of ssm for "Same-Sex Marriage Script by Gay Canvasser"
sd(ss_gay$ssm)

#Mean ssm for "No Contact"
no_contact <- s1w1[s1w1$treatment=="No Contact",]
mean(no_contact$ssm)

#Standard deviation of ssm for "No Contact"
sd(no_contact$ssm)

```

-   Mean ssm for "Same-Sex Marriage Script by Gay Canvasser": **3.03**

-   Standard deviation of ssm for "Same-Sex Marriage Script by Gay Canvasser": **1.67**

-   Mean ssm for "No Contact": **3.04**

-   Standard deviation of ssm for "No Contact": **1.68**

Standard mean difference = $\frac{\bar{x}_{gay canvasser} - \bar{x}_{no contact}}{\sqrt{(s^2_{gay canvasser} + s^2_{no contact})/2}}$

$\frac{3.025195 - 3.042764}{\sqrt{(1.666099)^2 + (1.68054)^2)/2}}$

$\frac{-0.017569}{1.673335078}$

Standardized difference in means: **-0.0105**

Since the SDM of -0.0105 has an absolute value less than 0.25, we can conclude that these groups look similar at baseline. Both the treatment group (same-sex marriage script by gay canvasser) and control group (no contact) have a similar baseline mean (3.03 and 3.04) and standard deviation (1.67 and 1.68). The standardized difference in means confirms with a single statistic that these baseline groups from wave 1 are very similar.

***Q1.6 (4 points):** Now look at the outcomes. Filter to Study 1, Wave 2 only. Calculate the mean ssm for the gay canvasser group and the No Contact group. What is the difference in means (gay canvasser minus no contact)?*

*Interpret: since the groups were balanced at baseline, what does a difference at Wave 2 suggest? (2–3 sentences)*

```{r}
#Subsetting to study 1, wave 1
s1w2 <- study1[study1$wave==2,]

#Mean ssm for "Same-Sex Marriage Script by Gay Canvasser"
wave2_ss_gay <- s1w2[s1w2$treatment=="Same-Sex Marriage Script by Gay Canvasser",]
mean(wave2_ss_gay$ssm)

#Mean ssm for "No Contact"
wave2_no_contact <- s1w2[s1w2$treatment=="No Contact",]
mean(wave2_no_contact$ssm)
```

Mean ssm for gay canvasser group: **3.139**

Mean ssm for no contact group: **3.04**

Difference in means: 3.139 - 3.04 = **0.1**

Given that these two groups were balanced at baseline, a difference at wave 2 suggests that the treatment of a gay canvasser reading a script about same-sex marriage led to an increase in the thermometer score (ssm) indicating a positive perception of same-sex marriage. Compared to the control group of "no contact" which saw no change in mean ssm, the treatment group saw nearly a 0.11-unit increase in ssm. Since the design is stated to be experimental, we would conclude from this that the treatment of a gay canvasser reading a script about same-sex marriage caused a positive increase in participants' sentiment toward same-sex marriage.

## Part 2: The Forensic Discovery

*Broockman, Kalla, and Aronow showed that LaCour fabricated his data — and they caught him using exactly the tools from this course: histograms, means, standard deviations, and correlations. You’ll now retrace their steps.*

**2.1 The Baseline Distribution**

***Q2.1 (3 points):** Load gayreshaped.csv and ccap2012.csv. Report the number of observations and column names for each dataset. In gayreshaped.csv, the variables therm1 through therm4 represent feeling thermometer scores (0–100) toward gay men and lesbians, measured at each survey wave.*

```{r}
#Loading datasets
#| warning: false
gay_r <- read_csv("./data/raw/gayreshaped.csv")
ccap <- read_csv("./data/raw/ccap2012.csv")

#

```

**gay_r**

-   Number of observations: \_\_

-   Column names: \_\_

**ccap**

-   Number of observations: \_\_

-   Column names: \_\_

**Q2.2 (5 points):** Extract the baseline feeling thermometer (therm1) for Study 1 from gayreshaped.csv. Calculate its mean, median, and standard deviation. Then do the same for gaytherm from ccap2012.csv (remove NAs). Report both sets of statistics side by side. What do you notice? (1–2 sentences)

```{r}

```

*What do you notice?*

Here

***Q2.3 (5 points):** Create side-by-side histograms comparing the LaCour Study 1 baseline (therm1) and the CCAP data (gaytherm). Use the same bin breaks for both: breaks = seq(0, 100, by = 5).*

*Hint: Combine the two datasets into one data frame with a column indicating the source, then use ggplot() with facet_wrap().*

```{r}

```

*Do these look like they came from different populations? (2–3 sentences)*

Here

**2.2 Re-test Reliability**

*When real people take the same survey twice, their answers correlate — but not perfectly. People’s moods change, they round differently, they misread questions. A re-test correlation of 0.95–0.97 is normal for feeling thermometers. A correlation near 1.0 is suspicious.*

***Q2.4 (6 points):** For both the Study 1 and Study 2 control groups (“No Contact”), calculate:*

*(a) The correlation between therm1 and therm2 (use cor() with use = "complete.obs")*

*(b) The standard deviation of the within-person change (therm2 - therm1)*

*Present your results side by side. A re-test correlation of 0.95–0.97 is typical for real feeling thermometer data. Which study shows a more plausible pattern, and why? (3–4 sentences)*

```{r}

```

Answer here

***Q2.5 (5 points):** Create two scatter plots side by side:*

*(a) Study 1 control group: therm1 (x-axis) vs therm2 (y-axis), with a 45-degree line*

*(b) Study 2 control group: therm1 (x-axis) vs therm2 (y-axis), with a 45-degree line*

*Describe the differences. Which looks more like real longitudinal data, and why? (3–4 sentences)*

```{r}

```

Here

**2.3 Putting It All Together**

***Q2.6 (6 points):** You’ve now found three irregularities:*

*1. The baseline distributions match an existing national survey*

*2. The re-test correlations are implausibly high (Study 1)*

*3. The within-person changes are implausibly small (Study 1)*

*In 4–5 sentences, explain why these three findings together suggest the data were fabricated rather than collected in a real experiment. Use specific numbers from your analysis. Why would a fabricator produce data with these properties?*

Here

## Part 3: What Real Data Looks Like

*You’ve found irregularities in Study 1. Now analyze Study 2 and see if the same patterns hold.*

**3.1 Comparing Baselines**

***Q3.1 (3 points):** Filter gayreshaped.csv to Study 2. How many observations? Calculate the mean and SD of therm1 (baseline) for the treatment group and the control group separately.*

```{r}

```

Here

***Q3.2 (3 points):** Create side-by-side histograms of Study 1 baseline (therm1) and Study 2 baseline (therm1), using the same bins as before. How do they compare visually? (2–3 sentences)*

```{r}

```

Here

**3.2 Checking Balance and Comparing Outcomes in Study 2**

***Q3.3 (3 points):** Filter gayreshaped.csv to Study 2, keeping only the “Same-Sex Marriage Script by Gay Canvasser” and “No Contact” groups. Check balance at baseline: calculate the mean and SD of therm1 for each group and the standardized difference in means (same formula as Q1.5). Do the groups look balanced? (1–2 sentences)*

```{r}

```

Here

***Q3.4 (3 points):** Now compare outcomes. Calculate the mean therm2 (post-conversation) for each group. What is the difference in post-period means (gay canvasser minus no contact)?*

```{r}

```

**3.3 Uncertainty Around the Difference**

***Q3.5 (4 points):** Calculate the standard error of the difference in post-period means from Q3.4.Use the formula, where 𝑠2 is the variance of therm2 and 𝑛 is the sample size for each group. Then construct a 95% confidence interval: ( ̄𝑥1 − ̄𝑥2) ± 1.96 × 𝑆𝐸. Does the CI include zero? What does this tell you about whether the groups differ after the conversations?*

```{r}

```

Here

**3.4 Tracking Attitudes Over Time**

***Q3.6 (4 points):** For the Study 2 control group, calculate the correlation between therm1 and each subsequent wave: therm2, therm3, and therm4. Use cor() with use = "complete.obs".*

*Do the correlations decay over time (i.e., do more distant waves correlate less with baseline)? What does this pattern tell you about the stability of real survey measurements? (2–3 sentences)*

```{r}

```

Here

***Q3.7 (5 points):** You’ve now analyzed fabricated data (Study 1), caught the fraud, and examined more realistic data (Study 2). In 5–6 sentences, discuss:*

*• What makes Study 2’s data different from Study 1’s data? Be specific about which statistics differ and by how much.*

*• Why is it important that real treatment effects are smaller and messier than fabricated ones?*

*• What role did simple measurement tools (mean, SD, histograms, correlations) play in detecting the fraud?*

Here

## Part 4: Formalizing the Evidence with Regression

*Now formalize your findings using regression.*

**4.1 Regression as a Fraud Detector**

***Q4.1 (5 points):** For the Study 1 control group, fit a linear regression predicting therm2 from therm1:*

*therm2𝑖 = 𝛽0 + 𝛽1 × therm1𝑖 + 𝜖𝑖*

*Store the result as fit_s1 and display the summary. Report the slope (̂ 𝛽1), the R-squared, and the residual standard error.*

*Hint: Use lm(therm2 \~ therm1, data = ...) and summary().*

```{r}

```

***Q4.2 (5 points):** Repeat Q4.1 for the Study 2 control group. Store as fit_s2. Report the same three quantities: slope, R-squared, and residual standard error.*

*Compare the two models. In fabricated data, we’d expect̂ 𝛽1 ≈ 1 and 𝑅2 ≈ 1 because the “follow-up” data is just a copy of the baseline with tiny noise added. In real data, we’d expect regression to the mean:̂ 𝛽1 \< 1 and lower 𝑅2. Which pattern do you see? (3–4 sentences)*

```{r}

```

Here

**4.2 Comparing Groups with Regression**

***Q4.3 (5 points):** Now use regression to compare the two groups in Study 2. Create a binary variable treated that equals 1 for the “Same-Sex Marriage Script by Gay Canvasser” group and 0 for the “No Contact” group. Then fit:*

*therm2𝑖 = 𝛽0 + 𝛽1 × treated𝑖 + 𝜖𝑖*

*Report̂ 𝛽1, its standard error, t-statistic, and p-value. Comparê 𝛽1 to the difference in means you calculated in Q3.4 — what do you notice? What doeŝ 𝛽0 estimate? Is the difference between groups statistically significant at 𝛼 = 0.05?*

*Note: Filter to only the gay canvasser and no contact groups before running the regression.*

```{r}

```

Here

**4.3 Controlling for Baseline Attitudes**

***Q4.4 (5 points):** Add therm1 as a control variable:*

*therm2𝑖 = 𝛽0 + 𝛽1 × treated𝑖 + 𝛽2 × therm1𝑖 + 𝜖𝑖*

*Report the coeﬀicient on treated, its standard error, t-statistic, and p-value. Compare this to your results from Q4.3:*

*(a) Did the estimate of̂ 𝛽1 change much?*

*(b) Did the standard error change? In which direction?*

*(c) Why does controlling for baseline attitudes improve precision? (2–3 sentences)*

```{r}

```

Here

***Q4.5 (5 points):** Compare the R-squared values from Q4.3 and Q4.4. Why does adding therm1 increase the R-squared so dramatically? What does this tell you about how much of the variation in follow-up attitudes is explained by where people started versus the treatment?*

*Finally, use predict() to generate predicted values from your Q4.4 model for two hypothetical respondents:*

*• Person A: treated = 1, therm1 = 50 (moderate baseline)*

*• Person B: treated = 0, therm1 = 50 (same baseline, no treatment)*

*What is the predicted difference in their Wave 2 attitudes?*

```{r}

```

Here
